Best Practices in Latent Class Analysis using the Free Open Source Software

### 2023-04-11 manuscript.Rmd
Final Proofread pt.3
Minor adjustments to phrasing etc.

Particular points of attention include measurement level of the indicators.
Recall that `tidySEM` can estimate mixture models with either numeric (`numeric` or `integer`) or `ordered` type indicators.
The column `type` should thus only contain these types.
Also check that the number of `unique` values match the intended measurement level.
Numeric variables with few unique values are typically better modeled as ordinal.
Our experience consulting on LCA methods
and moderating the `tidyLPA` Google group suggest that the
misapplication of mixture models to ordinal (e.g., Likert-type)
indicators is the most common source of user error. Whereas it has been
argued that some parametric methods are robust when scales with 7+
response categories are treated as continuous [e.g., @norman_likert_2010; rhemtulla_when_2012],
it is very unlikely
that an ordinal variable can be modeled as a *mixture* of multiple
normal distributions.
Consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
Each class-specific mean would describe a single response
category,
and a class-specific variance component would be empirically underidentified.
In sum, variables with few unique values might not be suitable for Gaussian mixture modeling.
LCA with ordinal indicators is more appropriate.
<!-- Descriptive statistics and -->
<!-- plots can be relegated to online supplements, provided that these are -->
<!-- readily accessible [consider using a GitHub repository as a -->
<!-- comprehensive public research archive, as explained in -->
<!-- @van_lissa_worcs_2021]. -->

@@

The second consideration in model specification is the risk of overfitting [@hastie_elements_2009].
Overfitting occurs when a model has many parameters relative to the number of observations,
causing it to capture idiosyncratic noise in the sample and limiting its generalizability to new samples.
As explained before, the number of parameters in LCA scales with the number of estimated classes.
Consequently, LCA with many classes have a potentially very high number of parameters.
In Bayesian mixture modeling,
it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased [@rousseau_asymptotic_2011].
Although it is not known whether frequentist LCA exhibits the same behavior,
low class counts are still an indicator of potential overfitting.
The section on Class Enumeration addresses ways to guard against overfitting.

Models can be specified in `tidySEM` in different ways.
The wrapper functions `mx_profiles()` and `mx_lca()` construct LCA models for continuous and ordinal indicators respectively, without needing to specify model syntax.
The function `mx_profiles()` has two arguments to free elements of the covariance matrix: `variances` and `covariances`.
By default, `mx_profiles()` fixes covariances to zero, and thus performs latent profile analysis.
Finally, function `mx_mixture()` accepts a more flexible `model` argument, which can be a character string specifying a model in `lavaan` syntax,
which is translated to `OpenMx` models via the function `as_ram()`,
or a list of `OpenMx` models.

@@

One problem with the random starts approach is that it is
computationally expensive to run this many replications. Moreover,
because the algorithm begins with random starting values,
it is inefficient because many marbles are likely to start far away from any valley.
Additionally, if all sets of starting values are close to a specific local optimum,
there is a risk that the global optimum is never found.
An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing [SA, @corana_minimizing_1987].
SA does not behave quite like the marbles of the random starts approach;
instead, it iteratively considers some "destination" in the landscape,
and compares its $-2LL$ to the current one.
If the destination $-2LL$ is lower, 
the estimator moves there.
However, if the destination $-2LL$ is *higher*,
it still moves there occasionally, with a certain probability.
Think of this as occasionally flicking the marble,
to see if it rolls back into the current valley, or finds another,
even deeper one.
This property allows SA it to escape local optima and find the global optimum.

@@

Nonconvergence due to bad starting values
can be overcome by manually specifying starting values:
Specify the argument `run = FALSE` in the call to `mx_mixture()` and related functions.
This returns the model with default starting values, which can be updated.
Nonconvergence and local optima may also be overcome by re-estimating the model while permuting the starting values.
The `OpenMx` package contains a family of functions for this purpose: `mxTryHard()` and `mxTryHardWideSearch()` for continuous indicators, and `mxTryHardOrdinal()` for ordinal indicators.
When poor data quality is to blame, the only solution is to collect more, or higher quality, data.

@@

In class enumeration, the set of models is compared on several statistical and substantive criteria.
In exploratory LCA, these criteria may be decisive in selecting a final model;
in the confirmatory case, the burden of proof is reversed: there must be convincing evidence against the theoretical model in order to reject it.
Statistical criteria to consider in class enumeration include model fit indices and
likelihood ratio tests [@weller_latent_2020; @masyn_latent_2013].
Substantive criteria include convergence and model identification,
class separability,
and interpretability.
We address each of these criteria below.

@@

Due to the lack of absolute fit indices for LCA, the assessment of relative fit
relies on information criteria.
Information criteria (ICs) are computed from the $-2LL$,
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
In addition to capturing model fit, ICs add a penalty for model complexity thus
preferring more model fit and parsimony.
The lower the value of an information criterion, the better the overall fit of the
model.
The *Akaike Information Criterion* (*AIC*) is the original information criterion.
It is computed as $-2LL+2p$, where $p$ is the number of parameters.
The *Bayesian Information Criterion* (*BIC*) multiplies the penalty for complexity with a function of the sample size, $p \ln(n)$,
such that complex models are penalized more in smaller samples.
Finally,
the so-called *sample size adjusted BIC* (*saBIC*), $\ln(\frac{n + 2}{24})$,
implements a penalty for sample size that is based on the Minimum Description Length, a concept from data compression that refers to the smallest number of pieces of information necessary to reproduce the data [@Rissanen_1985; @sclove_application_1987].
Simulation studies have shown that the saBIC performs well for enumeration of homogenous clusters in small samples,
but tends to select too many classes when the number of the number of classes and the sample size are large [@tein_statistical_2013; @chen_efficacy_2017].
The BIC has been found to perform consistently well across simulation studies,
and is among the most widely known information criteria.
As such, BIC may be a sensible default criterion for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
In `tidySEM`, all three ICs are available as a part of the output of `table_fit()`.

@@

Given a specific information criterion, it is also possible to calculate so-called "IC-weights",
which indicate the relative support the data provides for each model in the set [@wagenmakers_aic_2004].
This is particularly useful for confirmatory LCA,
as it allows one to quantify the support for the theoretical model.
In `tidySEM`, IC-weights are calculated by calling `ic_weights()` on the model fit table.
Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.
<!-- The general objective of information criteria is to evaluate the model's  -->
<!-- out-of-sample predictive accuracy, -->
<!-- thus curtailing overfitting. -->
<!-- Fit measures such as $R^2$ evaluate the in-sample predictive accuracy,  -->
<!-- i.e. the model's ability to predict the observed outcomes based on the same  -->
<!-- data that was used to fit the model. In-sample metrics are positively biased,  -->
<!-- meaning that they overestimate the fit the model has in reality, i.e. when  -->
<!-- presented with new data. While common ways to overcome this bias is through -->
<!-- resampling methods such as cross-validation, or use of out-of-sample data such -->
<!-- as a test set, ICs correct for this positive bias by evaluating the model's  -->
<!-- accuracy in a way which approximates the out-of-sample predictive accuracy.  -->
<!-- ICs achieve this by applying different penalty metrics. For example, $R^2$ will  -->
<!-- increase even if an added predictor is fully redundant. ICs show preference for -->
<!-- simpler models and indicate worse fit when a predictor, or its added complexity  -->
<!-- is unnecessary [@mcelreath_statistical_2020].  -->

@@

Another solution is to derive an empirical sampling distribution for the LR difference.
The so-called *Bootstrapped Likelihood Ratio Test (BLRT)* does so by simulating data from the $k-1$-class model $b$ times (e.g., 1000),
and fitting the $k$- and $k-1$-class models to each simulated dataset [@nylund_deciding_2007]. 
The p-value of this test is the proportion of $b$ in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
Simulation studies suggest that the BLRT has greater statistical power than the LRT,
although both perform comparably well at detecting the true number of classes [@nylund_deciding_2007; @tein_statistical_2013].
A crucial limitation of the BLRT is that it is very computationally expensive.
With this in mind, `table_fit()` does not provide the BLRT by default.
Specific models can be compared using the function `BLRT()`.

>>>>

Particular points of attention include measurement level of the indicators.
Recall that `tidySEM` can estimate mixture models with either numeric (`numeric` or `integer`) or `ordered` type indicators.
The column `type` should thus only contain these types.
Also check that the number of `unique` values match the intended measurement level.
Numeric variables with few unique values are typically better modeled as ordinal.
Our experience consulting on LCA methods
and moderating the `tidyLPA` Google group suggest that the
misapplication of mixture models to ordinal (e.g., Likert-type)
indicators is the most common source of user error. Whereas it has been
argued that some parametric methods are robust when scales with 7+
response categories are treated as continuous [e.g., @norman_likert_2010; @rhemtulla_when_2012],
it is very unlikely
that an ordinal variable can be modeled as a *mixture* of multiple
normal distributions.
Consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
Each class-specific mean would describe a single response
category,
and a class-specific variance component would be empirically underidentified.
In sum, variables with few unique values might not be suitable for Gaussian mixture modeling.
LCA with ordinal indicators is more appropriate.
<!-- Descriptive statistics and -->
<!-- plots can be relegated to online supplements, provided that these are -->
<!-- readily accessible [consider using a GitHub repository as a -->
<!-- comprehensive public research archive, as explained in -->
<!-- @van_lissa_worcs_2021]. -->

@@

The second consideration in model specification is the risk of overfitting [@hastie_elements_2009].
Overfitting occurs when a model has many parameters relative to the number of observations,
causing it to capture idiosyncratic noise in the sample and limiting its generalizability to new samples.
As explained before, the number of parameters in LCA scales with the number of estimated classes.
Consequently, LCA with many classes have a large number of parameters.
In Bayesian mixture modeling,
it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased [@rousseau_asymptotic_2011].
Although it is not known whether frequentist LCA exhibits the same behavior,
low class counts are still an indicator of potential overfitting.
The section on Class Enumeration addresses ways to guard against overfitting.

In `tidySEM` models can be specified in different ways.
The wrapper functions `mx_profiles()` and `mx_lca()` construct LCA models for continuous and ordinal indicators respectively, without needing to specify model syntax.
The function `mx_profiles()` has two arguments to free elements of the covariance matrix: `variances` and `covariances`.
By default, `mx_profiles()` fixes covariances to zero, and thus performs latent profile analysis.
Finally, function `mx_mixture()` accepts a more flexible `model` argument, which can be a character string specifying a model in `lavaan` syntax,
which is translated to `OpenMx` models via the function `as_ram()`,
or a list of `OpenMx` models.

@@

One problem with the random starts approach is that it is
computationally expensive to run this many replications. Moreover,
because the algorithm begins with random starting values,
it is inefficient because many marbles are likely to start far away from any valley.
Additionally, if all sets of starting values are close to a specific local optimum,
there is a risk that the global optimum is never found.
An alternative solution that overcomes these challenges is to use a global optimizer, like *simulated annealing* [*SA*, @corana_minimizing_1987].
SA does not behave quite like the marbles of the random starts approach;
instead, it iteratively considers some "destination" in the landscape,
and compares its $-2LL$ to the current one.
If the destination $-2LL$ is lower, 
the estimator moves there.
However, if the destination $-2LL$ is *higher*,
it still moves there occasionally, with a certain probability.
Think of this as occasionally flicking the marble,
to see if it rolls back into the current valley, or finds another,
even deeper one.
This property allows SA it to escape local optima and find the global optimum.

@@

Nonconvergence due to bad starting values
can be overcome by manually specifying starting values:
Specify the argument `run = FALSE` in the call to `mx_mixture()` and related functions.
This returns the model with default starting values, which can be updated.
Nonconvergence and local optima may also be overcome by re-estimating the model while permuting the starting values.
The `OpenMx` package contains a family of functions for this purpose: `mxTryHard()` and `mxTryHardWideSearch()` for continuous indicators, and `mxTryHardOrdinal()` for ordinal indicators.
When poor data quality is to blame, the only solution is to collect more, or higher quality data.

@@

In class enumeration, the set of models is compared on several statistical and substantive criteria.
In exploratory LCA, these criteria may be decisive in selecting a final model;
in the confirmatory case, the burden of proof is reversed: there must be convincing evidence against the theoretical model in order to reject it.
Statistical criteria to consider in class enumeration include model fit indices and
likelihood ratio tests [@weller_latent_2020; @masyn_latent_2013].
Substantive criteria include model convergence and identification, and
class separability,
and interpretability.
We address each of these criteria below.

@@

Due to the lack of absolute fit indices for LCA, the assessment of relative fit
relies on information criteria.
*Information criteria* (*ICs*) are computed from the $-2LL$,
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
In addition to capturing model fit, ICs add a penalty for model complexity thus
preferring more model fit and parsimony.
The lower the value of an information criterion, the better the overall fit of the
model.
The *Akaike Information Criterion* (*AIC*) is the original information criterion.
It is computed as $-2LL+2p$, where $p$ is the number of parameters.
The *Bayesian Information Criterion* (*BIC*) multiplies the penalty for complexity with a function of the sample size, $p \ln(n)$,
such that complex models are penalized more in smaller samples.
Finally,
the so-called *sample size adjusted BIC* (*saBIC*), $\ln(\frac{n + 2}{24})$,
implements a penalty for sample size that is based on the Minimum Description Length, a concept from data compression that refers to the smallest number of pieces of information necessary to reproduce the data [@Rissanen_1985; @sclove_application_1987].
Simulation studies have shown that the saBIC performs well for enumeration of homogenous clusters in small samples,
but tends to select too many classes when the number of the number of classes and the sample size are large [@tein_statistical_2013; @chen_efficacy_2017].
The BIC has been found to perform consistently well across simulation studies,
and is among the most widely known information criteria.
As such, BIC may be a sensible default criterion for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
In `tidySEM`, all three ICs are available as a part of the output of `table_fit()`.

@@

Given a specific information criterion, it is also possible to calculate *IC-weights*,
which indicate the relative support the data provides for each model under consideration [@wagenmakers_aic_2004].
This is particularly useful for confirmatory LCA,
as it allows one to quantify the support for the theoretical model.
In `tidySEM`, IC-weights are calculated by calling `ic_weights()` on the model fit table.
Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.
<!-- The general objective of information criteria is to evaluate the model's  -->
<!-- out-of-sample predictive accuracy, -->
<!-- thus curtailing overfitting. -->
<!-- Fit measures such as $R^2$ evaluate the in-sample predictive accuracy,  -->
<!-- i.e. the model's ability to predict the observed outcomes based on the same  -->
<!-- data that was used to fit the model. In-sample metrics are positively biased,  -->
<!-- meaning that they overestimate the fit the model has in reality, i.e. when  -->
<!-- presented with new data. While common ways to overcome this bias is through -->
<!-- resampling methods such as cross-validation, or use of out-of-sample data such -->
<!-- as a test set, ICs correct for this positive bias by evaluating the model's  -->
<!-- accuracy in a way which approximates the out-of-sample predictive accuracy.  -->
<!-- ICs achieve this by applying different penalty metrics. For example, $R^2$ will  -->
<!-- increase even if an added predictor is fully redundant. ICs show preference for -->
<!-- simpler models and indicate worse fit when a predictor, or its added complexity  -->
<!-- is unnecessary [@mcelreath_statistical_2020].  -->

@@

Another solution is to derive an empirical sampling distribution for the LR difference.
The so-called *Bootstrapped Likelihood Ratio Test (BLRT)* does so by simulating data from the $k-1$-class model $b$ times (e.g., 1000),
and fitting the $k$- and $k-1$-class models to each simulated dataset [@nylund_deciding_2007]. 
The p-value of this test is the proportion of $b$ in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
Simulation studies suggest that the BLRT has greater statistical power than the LRT,
although both perform comparably well at detecting the true number of classes [@nylund_deciding_2007; @tein_statistical_2013].
A crucial limitation of the BLRT is that it is very computationally expensive.
With this in mind, `table_fit()` does not provide the BLRT by default.
Instead, specific models can be compared using the function `BLRT()`.

### 2023-04-11 manuscript.tex
Final Proofread pt.3
Minor adjustments to phrasing etc.

Particular points of attention include measurement level of the indicators.
Recall that \texttt{tidySEM} can estimate mixture models with either numeric (\texttt{numeric} or \texttt{integer}) or \texttt{ordered} type indicators.
The column \texttt{type} should thus only contain these types.
Also check that the number of \texttt{unique} values match the intended measurement level.
Numeric variables with few unique values are typically better modeled as ordinal.
Our experience consulting on LCA methods
and moderating the \texttt{tidyLPA} Google group suggest that the
misapplication of mixture models to ordinal (e.g., Likert-type)
indicators is the most common source of user error. Whereas it has been
argued that some parametric methods are robust when scales with 7+
response categories are treated as continuous {[}e.g., Norman (2010); rhemtulla\_when\_2012{]},
it is very unlikely
that an ordinal variable can be modeled as a \emph{mixture} of multiple
normal distributions.
Consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
Each class-specific mean would describe a single response
category,
and a class-specific variance component would be empirically underidentified.
In sum, variables with few unique values might not be suitable for Gaussian mixture modeling.
LCA with ordinal indicators is more appropriate.

@@

The second consideration in model specification is the risk of overfitting (Hastie, Tibshirani, \& Friedman, 2009).
Overfitting occurs when a model has many parameters relative to the number of observations,
causing it to capture idiosyncratic noise in the sample and limiting its generalizability to new samples.
As explained before, the number of parameters in LCA scales with the number of estimated classes.
Consequently, LCA with many classes have a potentially very high number of parameters.
In Bayesian mixture modeling,
it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased (Rousseau \& Mengersen, 2011).
Although it is not known whether frequentist LCA exhibits the same behavior,
low class counts are still an indicator of potential overfitting.
The section on Class Enumeration addresses ways to guard against overfitting.

Models can be specified in \texttt{tidySEM} in different ways.
The wrapper functions \texttt{mx\_profiles()} and \texttt{mx\_lca()} construct LCA models for continuous and ordinal indicators respectively, without needing to specify model syntax.
The function \texttt{mx\_profiles()} has two arguments to free elements of the covariance matrix: \texttt{variances} and \texttt{covariances}.
By default, \texttt{mx\_profiles()} fixes covariances to zero, and thus performs latent profile analysis.
Finally, function \texttt{mx\_mixture()} accepts a more flexible \texttt{model} argument, which can be a character string specifying a model in \texttt{lavaan} syntax,
which is translated to \texttt{OpenMx} models via the function \texttt{as\_ram()},
or a list of \texttt{OpenMx} models.

@@

One problem with the random starts approach is that it is
computationally expensive to run this many replications. Moreover,
because the algorithm begins with random starting values,
it is inefficient because many marbles are likely to start far away from any valley.
Additionally, if all sets of starting values are close to a specific local optimum,
there is a risk that the global optimum is never found.
An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing (SA, Corana et al., 1987).
SA does not behave quite like the marbles of the random starts approach;
instead, it iteratively considers some ``destination'' in the landscape,
and compares its \(-2LL\) to the current one.
If the destination \(-2LL\) is lower,
the estimator moves there.
However, if the destination \(-2LL\) is \emph{higher},
it still moves there occasionally, with a certain probability.
Think of this as occasionally flicking the marble,
to see if it rolls back into the current valley, or finds another,
even deeper one.
This property allows SA it to escape local optima and find the global optimum.

@@

Nonconvergence due to bad starting values
can be overcome by manually specifying starting values:
Specify the argument \texttt{run\ =\ FALSE} in the call to \texttt{mx\_mixture()} and related functions.
This returns the model with default starting values, which can be updated.
Nonconvergence and local optima may also be overcome by re-estimating the model while permuting the starting values.
The \texttt{OpenMx} package contains a family of functions for this purpose: \texttt{mxTryHard()} and \texttt{mxTryHardWideSearch()} for continuous indicators, and \texttt{mxTryHardOrdinal()} for ordinal indicators.
When poor data quality is to blame, the only solution is to collect more, or higher quality, data.

@@

In class enumeration, the set of models is compared on several statistical and substantive criteria.
In exploratory LCA, these criteria may be decisive in selecting a final model;
in the confirmatory case, the burden of proof is reversed: there must be convincing evidence against the theoretical model in order to reject it.
Statistical criteria to consider in class enumeration include model fit indices and
likelihood ratio tests (Masyn, 2013; Weller et al., 2020).
Substantive criteria include convergence and model identification,
class separability,
and interpretability.
We address each of these criteria below.

@@

Due to the lack of absolute fit indices for LCA, the assessment of relative fit
relies on information criteria.
Information criteria (ICs) are computed from the \(-2LL\),
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
In addition to capturing model fit, ICs add a penalty for model complexity thus
preferring more model fit and parsimony.
The lower the value of an information criterion, the better the overall fit of the
model.
The \emph{Akaike Information Criterion} (\emph{AIC}) is the original information criterion.
It is computed as \(-2LL+2p\), where \(p\) is the number of parameters.
The \emph{Bayesian Information Criterion} (\emph{BIC}) multiplies the penalty for complexity with a function of the sample size, \(p \ln(n)\),
such that complex models are penalized more in smaller samples.
Finally,
the so-called \emph{sample size adjusted BIC} (\emph{saBIC}), \(\ln(\frac{n + 2}{24})\),
implements a penalty for sample size that is based on the Minimum Description Length, a concept from data compression that refers to the smallest number of pieces of information necessary to reproduce the data (Rissanen, 1983; Sclove, 1987).
Simulation studies have shown that the saBIC performs well for enumeration of homogenous clusters in small samples,
but tends to select too many classes when the number of the number of classes and the sample size are large (Chen, Luo, Palardy, Glaman, \& McEnturff, 2017; Tein, Coxe, \& Cham, 2013).
The BIC has been found to perform consistently well across simulation studies,
and is among the most widely known information criteria.
As such, BIC may be a sensible default criterion for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
In \texttt{tidySEM}, all three ICs are available as a part of the output of \texttt{table\_fit()}.

@@

Given a specific information criterion, it is also possible to calculate so-called ``IC-weights'',
which indicate the relative support the data provides for each model in the set (Wagenmakers \& Farrell, 2004).
This is particularly useful for confirmatory LCA,
as it allows one to quantify the support for the theoretical model.
In \texttt{tidySEM}, IC-weights are calculated by calling \texttt{ic\_weights()} on the model fit table.
Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.

@@

Another solution is to derive an empirical sampling distribution for the LR difference.
The so-called \emph{Bootstrapped Likelihood Ratio Test (BLRT)} does so by simulating data from the \(k-1\)-class model \(b\) times (e.g., 1000),
and fitting the \(k\)- and \(k-1\)-class models to each simulated dataset (Nylund et al., 2007).
The p-value of this test is the proportion of \(b\) in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
Simulation studies suggest that the BLRT has greater statistical power than the LRT,
although both perform comparably well at detecting the true number of classes (Nylund et al., 2007; Tein et al., 2013).
A crucial limitation of the BLRT is that it is very computationally expensive.
With this in mind, \texttt{table\_fit()} does not provide the BLRT by default.
Specific models can be compared using the function \texttt{BLRT()}.

@@

>>>>

Particular points of attention include measurement level of the indicators.
Recall that \texttt{tidySEM} can estimate mixture models with either numeric (\texttt{numeric} or \texttt{integer}) or \texttt{ordered} type indicators.
The column \texttt{type} should thus only contain these types.
Also check that the number of \texttt{unique} values match the intended measurement level.
Numeric variables with few unique values are typically better modeled as ordinal.
Our experience consulting on LCA methods
and moderating the \texttt{tidyLPA} Google group suggest that the
misapplication of mixture models to ordinal (e.g., Likert-type)
indicators is the most common source of user error. Whereas it has been
argued that some parametric methods are robust when scales with 7+
response categories are treated as continuous (e.g., Norman, 2010; Rhemtulla, Brosseau-Liard, \& Savalei, 2012),
it is very unlikely
that an ordinal variable can be modeled as a \emph{mixture} of multiple
normal distributions.
Consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
Each class-specific mean would describe a single response
category,
and a class-specific variance component would be empirically underidentified.
In sum, variables with few unique values might not be suitable for Gaussian mixture modeling.
LCA with ordinal indicators is more appropriate.

@@

The second consideration in model specification is the risk of overfitting (Hastie, Tibshirani, \& Friedman, 2009).
Overfitting occurs when a model has many parameters relative to the number of observations,
causing it to capture idiosyncratic noise in the sample and limiting its generalizability to new samples.
As explained before, the number of parameters in LCA scales with the number of estimated classes.
Consequently, LCA with many classes have a large number of parameters.
In Bayesian mixture modeling,
it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased (Rousseau \& Mengersen, 2011).
Although it is not known whether frequentist LCA exhibits the same behavior,
low class counts are still an indicator of potential overfitting.
The section on Class Enumeration addresses ways to guard against overfitting.

In \texttt{tidySEM} models can be specified in different ways.
The wrapper functions \texttt{mx\_profiles()} and \texttt{mx\_lca()} construct LCA models for continuous and ordinal indicators respectively, without needing to specify model syntax.
The function \texttt{mx\_profiles()} has two arguments to free elements of the covariance matrix: \texttt{variances} and \texttt{covariances}.
By default, \texttt{mx\_profiles()} fixes covariances to zero, and thus performs latent profile analysis.
Finally, function \texttt{mx\_mixture()} accepts a more flexible \texttt{model} argument, which can be a character string specifying a model in \texttt{lavaan} syntax,
which is translated to \texttt{OpenMx} models via the function \texttt{as\_ram()},
or a list of \texttt{OpenMx} models.

@@

One problem with the random starts approach is that it is
computationally expensive to run this many replications. Moreover,
because the algorithm begins with random starting values,
it is inefficient because many marbles are likely to start far away from any valley.
Additionally, if all sets of starting values are close to a specific local optimum,
there is a risk that the global optimum is never found.
An alternative solution that overcomes these challenges is to use a global optimizer, like \emph{simulated annealing} (\emph{SA}, Corana et al., 1987).
SA does not behave quite like the marbles of the random starts approach;
instead, it iteratively considers some ``destination'' in the landscape,
and compares its \(-2LL\) to the current one.
If the destination \(-2LL\) is lower,
the estimator moves there.
However, if the destination \(-2LL\) is \emph{higher},
it still moves there occasionally, with a certain probability.
Think of this as occasionally flicking the marble,
to see if it rolls back into the current valley, or finds another,
even deeper one.
This property allows SA it to escape local optima and find the global optimum.

@@

Nonconvergence due to bad starting values
can be overcome by manually specifying starting values:
Specify the argument \texttt{run\ =\ FALSE} in the call to \texttt{mx\_mixture()} and related functions.
This returns the model with default starting values, which can be updated.
Nonconvergence and local optima may also be overcome by re-estimating the model while permuting the starting values.
The \texttt{OpenMx} package contains a family of functions for this purpose: \texttt{mxTryHard()} and \texttt{mxTryHardWideSearch()} for continuous indicators, and \texttt{mxTryHardOrdinal()} for ordinal indicators.
When poor data quality is to blame, the only solution is to collect more, or higher quality data.

@@

In class enumeration, the set of models is compared on several statistical and substantive criteria.
In exploratory LCA, these criteria may be decisive in selecting a final model;
in the confirmatory case, the burden of proof is reversed: there must be convincing evidence against the theoretical model in order to reject it.
Statistical criteria to consider in class enumeration include model fit indices and
likelihood ratio tests (Masyn, 2013; Weller et al., 2020).
Substantive criteria include model convergence and identification, and
class separability,
and interpretability.
We address each of these criteria below.

@@

Due to the lack of absolute fit indices for LCA, the assessment of relative fit
relies on information criteria.
\emph{Information criteria} (\emph{ICs}) are computed from the \(-2LL\),
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
In addition to capturing model fit, ICs add a penalty for model complexity thus
preferring more model fit and parsimony.
The lower the value of an information criterion, the better the overall fit of the
model.
The \emph{Akaike Information Criterion} (\emph{AIC}) is the original information criterion.
It is computed as \(-2LL+2p\), where \(p\) is the number of parameters.
The \emph{Bayesian Information Criterion} (\emph{BIC}) multiplies the penalty for complexity with a function of the sample size, \(p \ln(n)\),
such that complex models are penalized more in smaller samples.
Finally,
the so-called \emph{sample size adjusted BIC} (\emph{saBIC}), \(\ln(\frac{n + 2}{24})\),
implements a penalty for sample size that is based on the Minimum Description Length, a concept from data compression that refers to the smallest number of pieces of information necessary to reproduce the data (Rissanen, 1983; Sclove, 1987).
Simulation studies have shown that the saBIC performs well for enumeration of homogenous clusters in small samples,
but tends to select too many classes when the number of the number of classes and the sample size are large (Chen, Luo, Palardy, Glaman, \& McEnturff, 2017; Tein, Coxe, \& Cham, 2013).
The BIC has been found to perform consistently well across simulation studies,
and is among the most widely known information criteria.
As such, BIC may be a sensible default criterion for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
In \texttt{tidySEM}, all three ICs are available as a part of the output of \texttt{table\_fit()}.

@@

Given a specific information criterion, it is also possible to calculate \emph{IC-weights},
which indicate the relative support the data provides for each model under consideration (Wagenmakers \& Farrell, 2004).
This is particularly useful for confirmatory LCA,
as it allows one to quantify the support for the theoretical model.
In \texttt{tidySEM}, IC-weights are calculated by calling \texttt{ic\_weights()} on the model fit table.
Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.

@@

Another solution is to derive an empirical sampling distribution for the LR difference.
The so-called \emph{Bootstrapped Likelihood Ratio Test (BLRT)} does so by simulating data from the \(k-1\)-class model \(b\) times (e.g., 1000),
and fitting the \(k\)- and \(k-1\)-class models to each simulated dataset (Nylund et al., 2007).
The p-value of this test is the proportion of \(b\) in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
Simulation studies suggest that the BLRT has greater statistical power than the LRT,
although both perform comparably well at detecting the true number of classes (Nylund et al., 2007; Tein et al., 2013).
A crucial limitation of the BLRT is that it is very computationally expensive.
With this in mind, \texttt{table\_fit()} does not provide the BLRT by default.
Instead, specific models can be compared using the function \texttt{BLRT()}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-rhemtulla_when_2012}{}}%
Rhemtulla, M., Brosseau-Liard, P. É., \& Savalei, V. (2012). When can categorical variables be treated as continuous? A comparison of robust continuous and categorical {SEM} estimation methods under suboptimal conditions. \emph{Psychological Methods}, \emph{17}, 354--373. \url{https://doi.org/10.1037/a0029315}
